Artificial Intelligence for the Automated Synthesis and Validation of Programs

### 2018-09-13 doc/programming-projects/conselleria-emergente-18/AI-programming.tex
first full version project proposal

The performance of our AI method for the synthesis and validation of programs will be evaluated with regard to (1) {\bf computation time} and (2), {\bf memory}. Further, the programming and validation tasks used for the evaluation of the performance of our approach are of two different kinds:
\begin{itemize}
\item {\em Theoretical benchmarks}: Classic programming tasks are a neat touchstone to assess the performance of our approach. For instance, programs for the computation of mathematical/logic series, string manipulation and for the management of data structures such as {\em lists}, {\em queues}, {\em stacks} or {\em trees}. Figure~\ref{fig:sum} shows a synthesized program for computing $y = \sum_0^N x$. The program is pictured as a {\em finite state machine}: Machine nodes mount to the different program lines while edges are tagged with a {\em condition/instruction} label, that denotes the condition (over the program variables) under which program instructions are taken.  The program in Figure~\ref{fig:sum} assumes that the program variables are ${\mathcal V}=\{x,y,x=N,y=N\}$ and that the value of $x$ and $y$ is initially 0. The instruction set is ${\mathcal A}=\{x:=x+1, y:=y+1, x:=x+y, y:=y+x\}$. For the synthesis of this program we are using these five test cases $\{0 = \sum_0^0 x, 1 = \sum_0^1 x, 3 = \sum_0^2 x, 6 = \sum_0^3 x, 10 = \sum_0^4 x\}$ so the domains of variables $x$ and $y$ are ${\mathcal D}_x = {\mathcal D}_y = [0,10]$. Note that $x=N$ and $y=N$ are Boolean variables whose value depends on the current value of $x$ and $y$.

@@

\item {\em Real-world benchmarks}: We are involved in the four-year research project TIN2017-88476-C2-1-R from the {\em Spanish national plan} in which {\em AI Planning} and {\em activity recognition} is applied to different real-world domains such as {\em domotics}, {\em tourism}, {\em traffic control} and {\em robotics}. Interestingly many activities from these domains can be understood as simple programs. For instance, Figure~\ref{fig:activity} shows a five-line program (pictured as a finite state machine) that represents the sequence of instructions required for {\em making an orange juice}. We plan to evaluate our approach in synthesis and validation tasks coming from these real-world domains.

\begin{figure}[hbt!]
  \begin{center}
      \begin{scriptsize}   
    \begin{tikzpicture}[->,>=stealth',shorten >=1pt,auto,node distance=2.8cm,semithick]
	  \node[state] (A)              {$Q_0$};
	  \node[state] (B) [right of=A] {$Q_1$};
	  \node[state] (C) [right of=B] {$Q_2$};
	  \node[state] (D) [right of=C] {$Q_3$};
	  \node[state] (E) [right of=D] {$Q_4$};
	  \node[state] (F) [below of=E] {$Q_5$};                    
	  \path
          (A) edge [align=center] node { - /\\ $\mathsf{take}(orange)$} (B)          
          (B) edge [align=center] node { - /\\$\mathsf{take}(knife)$} (C)
          (C) edge [align=center] node { - /\\ $\mathsf{split}(orange)$} (D)
          (D) edge [align=center] node { - /\\ $\mathsf{squeeze}(orange)$} (E)
          (E) edge [align=center] node { ready(juice) /\\ - } (F)
          (E) edge [align=center, bend left] node { $\neg ready(juice)$ /\\ - } (A);                    
    \end{tikzpicture}
     \end{scriptsize}            
\end{center}
\caption{\small Five-line program representing the {\em making an orange juice} activity from the {\em domotics} domain.}
\label{fig:activity}
\end{figure}
\end{itemize}

\subsection{Specific objectives}
\label{sec:objectivos}
We characterize the programs objects of study in this project by these three dimensions:
\begin{itemize}
\item Number of {\em program lines}.
\item Size of the available {\em instruction set}.
\item Number and domain of {\em observable} program variables. That is, the subset of program variables whose value can condition the program flow.
\end{itemize}

To illustrate this, the program of Figure~\ref{fig:sum} for computing $y = \sum_0^N x$ have three program lines, an instruction set that comprises four instructions and a single observable Boolean variable ($x=N$). Likewise, the program of Figure~\ref{fig:activity}, representing the activity of {\em preparing an orange juice} have five program lines, an instruction set with four instructions (namely $\mathsf{take}(orange)$, $\mathsf{take}(knife)$, $\mathsf{split}(orange)$ and $\mathsf{squeeze}(orange)$) and a single observable Boolean variable, ${\tt ready(juice)}$, that holds when the orange juice is recognized to be ready.

The objective of this project is to study the performance of our approach (in terms of  {\bf computation time} and {\bf memory}) in the synthesis and validation of {\em TDD programming tasks} that come from the two kinds of domains described in the previous subsection. This objective can be itemized into two specific objectives:
\begin{enumerate}
\item To study the performance of our Artificial Intelligence approach for the {\bf synthesis of programs} up to: {\bf 10 program lines}, {\bf 10 observable variables} with binary domain and, an instruction set that comprises {\bf 10 instructions}. 
\item To study the performance of our Artificial Intelligence approach for the {\bf validation of programs} up to: {\bf 15 program lines}, {\bf 15 observable variables} with binary domain and, an instruction set that comprises {\bf 15 instructions}. 
\end{enumerate}

@@

We designed a 24-month workplan plan for the {\bf development of a user-interactive program synthesizer} that:
\begin{enumerate}
\item Takes as input a set of test cases that specify the TDD programming task to solve and outputs a program source code that passes these tests with a bug-free guarantee.
\item The {\em program synthesizer} must also work in {\em validation a mode}. In this particular setting, the {\em program synthesizer} receives an additional input specifying the program source code and it outputs a {\em validation certificate} guaranteeing that the input program passes (or fails) at the given test cases. 
\end{enumerate}

Figure~\ref{fig:gantt} details the proposed 24-month timeline for the project. A deliverable is provided at the end of each task.

@@

  \item {\bf T2. Development of the system architecture (months 7-12})
    \begin{small}
      \begin{enumerate}
      \item The programming-into-planning compiler ({\em Compiler 1}). This system component parses the {\em TDD programming task} and produces an {\em AIP task} encoded in the standard planning language PDDL.
      \item The plan-into-program compiler ({\em Compiler 2}). This system component extracts the program code and the corresponding validation certificate from the solution plan produced by an off-the-shelf AI planner.
      \end{enumerate}
\end{small}

@@

\item {\bf T3. Experiments and dissemination of results (months 13-24}).
   \begin{small}
      \begin{enumerate}
      \item Reporting the experimental performance of our AIP approach for solving diverse TDD programming tasks. This task will follow an iterative workflow over the following subtasks:
      \begin{enumerate}
      \item Executing the system architecture in the {\em theoretical benchmarks} described in Section~\ref{sec:evaluation}.
      \item Analysis and validation of the obtained results.
      \item Tuning and repairing the system components, evaluation metrics and benchmarks according to the obtained results.                 
      \item Executing the system architecture in the {\em real-world benchmarks} introduced in Section~\ref{sec:evaluation}.
      \item Analysis and validation of the obtained results.         
      \item Tuning and repairing the system components, evaluation metrics and benchmarks according to the obtained results.                 
      \end{enumerate}
      \item Dissemination of the obtained theoretical and empirical results by submitting papers to top international conferences and journals in AI.        
      \end{enumerate}
\end{small}        
{\small{\bf\em  Deliverable T3:} Final report with the obtained conclusions and produced publications.}
\end{enumerate}

\subsection{Budget}
Here we show the detailed desctiption of the two-year budget for the development of the proyect.

\begin{table}[hbt!]
\begin{small}  
  \begin{tabular}{clc|l}
    {\bf Priority} & {\bf Description} & {\bf Term} & {\bf Euros} \\\hline
1 & {\scriptsize Difusión de las actividades del grupo} & 2018, 2019 & 2,000\\    
2 & {\scriptsize Viajes, manutención y alojamiento grupo de investigación} & 2018 & 1,500\\
3 & {\scriptsize Viajes, manutención y alojamiento grupo de investigación} & 2019 & 1,500\\
4 & {\scriptsize Viajes, manutención, alojamiento y ponencias investigadores invitados} & 2018 & 1,500\\
5 & {\scriptsize Viajes, manutención, alojamiento y ponencias investigadores invitados} & 2019 & 1,500\\\hline
\multicolumn{3}{l|}{} & 8,000 \\
  \end{tabular}
\end{small}
\caption{\small Two-year budget for the development of the proyect.}
\end{table}

\section{Disemination and exploitation of the research results.}
\label{subsec:beneficios}
AIP has recently shown successful in {\em program testing} to generate {\em attack plans} that completed non-trivial software security tests~\cite{hoffmann2015simulated,steinmetz2016revisiting,shmaryahu2016constructing,steinmetz2016goal}. Promising research opportunities come from the application of AIP to {\em program synthesis} given that, {\em program synthesis} with a tests base, can be seen as the {\em program testing} dual.

In fact, our current work on {\em program synthesis} with AIP already produced several {\bf publications at top international conferences and journals on Artificial Intelligence}~\cite{segovia2017generating,sergio:aprogramingb:ijcai16,sergio:aprograming:ijcai16,sergio:aprograming:icaps16,segovia:FSC:JAIR2018,segovia:programs:AIJ19} and is the recipient of the {\it 2016 distinguished paper award} at the International Joint Conference on Artificial Intelligence, the main international conference on {\em Artificial intelligence}.

The main benefit of this project is to provide new insights into the current understanding of how AI can assist programmers in the software development.  Automated program synnthesis can produce huge cost savings in comparison to conventional programmming and increase the quality, trustbility and modifiability of programs. A longer term objective is the deeper and deeper embedding of intelligence in all manner of software systems, as well as the design and management of organizational processes by automated means.

In more detail, the expected benefits for this particular research project are four-fold:
\begin{enumerate}
\item A new {\bf evaluation methodology for the {\em Synthesis and Validation of Programs}}. The application of the exiting AIP technology to program synthesis can provide new evaluation metrics that assess how well a program covers a set of {\em input-output} test cases.  
\item An empirical {\bf study on the performance of the state-of-the-art AI planners for the {\em Synthesis and Validation of Programs}}. Research in AI algorithms is too often tested with laboratory problems and AIP is not an exception. Most of the new planning algorithms are only tested within the benchmarks of the International Planing Competition~\cite{vallati:IPC:AI15}. This project will help to meet the computational and expressiveness limits of AI planners when addressing real-world programming tasks. 
\item The {\bf development of open software and open benchmarks for the {\em Synthesis and Validation of Programs}}. We strongly belief that reproducibility and open knowledge are essential to the advance of the research on computer science. With this regard, we plan to develop a {\em github} repository where we make available the developed source code and benchmarks.
\item {\bf International dissemination of the obtained scientific results}. The scientific results obtained during the development of the project will be submitted to top Artificial Intelligence conferences (such as IJCAI, AAAI, ICML and ICAPS) and to the main journals in the AI field (such as AIJ, JMLR and JAIR).
\end{enumerate}

>>>>

The performance of our AI method for the synthesis and validation of programs will be evaluated with regard to (1), {\bf computation time} and (2), {\bf memory}. Further, the evaluation will be carried out in programming and validation tasks of two different kinds:
\begin{itemize}
\item {\bf\em Theoretical benchmarks}: Classic programming tasks are a neat touchstone to assess the performance of our approach. For instance, programs for the computation of mathematical/logic series, string manipulation and for the management of data structures such as {\em lists}, {\em queues}, {\em stacks} or {\em trees}. As an example Figure~\ref{fig:sum} shows a synthesized program, pictured as a {\em finite state machine}, for computing $y = \sum_0^N x$. The machine nodes mount to the different program lines while edges are tagged with a {\em condition/instruction} label, that denotes the condition (over the program variables) under which program instructions are taken.  The program in Figure~\ref{fig:sum} assumes that the program variables are ${\mathcal V}=\{x,y,x=N,y=N\}$ and that the value of $x$ and $y$ is initially 0. The instruction set is ${\mathcal A}=\{x:=x+1, y:=y+1, x:=x+y, y:=y+x\}$. For the synthesis of this program we are using the following five test cases $\{0 = \sum_0^0 x, 1 = \sum_0^1 x, 3 = \sum_0^2 x, 6 = \sum_0^3 x, 10 = \sum_0^4 x\}$ so the domains of variables $x$ and $y$ are ${\mathcal D}_x = {\mathcal D}_y = [0,10]$. Note that $x=N$ and $y=N$ are Boolean variables whose value depends on the current value of $x$ and $y$.

@@

\item {\bf\em Real-world benchmarks}: We are involved in the four-year research project {\sc ARPIA} (TIN2017-88476-C2-1-R, {\tt\small\url{http://arpia.blogs.upv.es/}}) fund by the {\em Spanish national plan} in which {\em AI Planning} and {\em activity recognition} is applied to different real-world domains such as {\em domotics}, {\em tourism}, {\em traffic control} and {\em robotics}. Interestingly many activities that are objects of study in these domains can be modeled as simple programs. Figure~\ref{fig:activity} shows a five-line program (pictured as a finite state machine) that represents the sequence of instructions required for {\em making an orange juice} from the {\em domotics} domain. We plan to evaluate our approach in synthesis and validation tasks coming from these real-world domains.

\begin{figure}[hbt!]
  \begin{center}
      \begin{scriptsize}   
    \begin{tikzpicture}[->,>=stealth',shorten >=1pt,auto,node distance=2.7cm,semithick]
	  \node[state] (A)              {$Q_0$};
	  \node[state] (B) [right of=A] {$Q_1$};
	  \node[state] (C) [right of=B] {$Q_2$};
	  \node[state] (D) [right of=C] {$Q_3$};
	  \node[state] (E) [right of=D] {$Q_4$};
	  \node[state] (F) [below of=E] {$Q_5$};                    
	  \path
          (A) edge [align=center] node { - /\\ $\mathsf{take}(orange)$} (B)          
          (B) edge [align=center] node { - /\\$\mathsf{take}(knife)$} (C)
          (C) edge [align=center] node { - /\\ $\mathsf{split}(orange)$} (D)
          (D) edge [align=center] node { - /\\ $\mathsf{squeeze}(orange)$} (E)
          (E) edge [align=center] node { ready(juice) /\\ - } (F)
          (E) edge [align=center, bend left] node {$\neg ready(juice)$ /\\ $\mathsf{take}(orange)$} (B);                    
    \end{tikzpicture}
     \end{scriptsize}            
\end{center}
\caption{\small Five-line program representing the {\em making an orange juice} activity from the {\em domotics} domain.}
\label{fig:activity}
\end{figure}
\end{itemize}

\subsection{Specific objectives}
\label{sec:objectivos}
We characterize the programs that are objects of study in this project by these three dimensions:
\begin{itemize}
\item Number of {\bf\em program lines}.
\item Size of the available {\bf\em instruction set}.
\item Number and domain of {\bf\em observable program variables}. That is, the subset of program variables whose value can condition the program flow.
\end{itemize}

To illustrate this, the program of Figure~\ref{fig:sum} for computing $y = \sum_0^N x$ have three program lines, an instruction set that comprises four instructions and a single observable Boolean variable ($x=N$). Likewise, the program of Figure~\ref{fig:activity}, representing the activity of {\em making an orange juice} have five program lines, an instruction set with four instructions (namely $\mathsf{take}(orange)$, $\mathsf{take}(knife)$, $\mathsf{split}(orange)$ and $\mathsf{squeeze}(orange)$) and the single observable Boolean variable, ${\tt ready(juice)}$, that holds when the orange juice is recognized to be ready.

The objective of this project is to study the performance of our approach (in terms of  {\em computation time} and {\em memory}) in the synthesis and validation of {\em TDD programming tasks} that come from domains of the two kinds described in the previous subsection. This general objective is itemized into two specific objectives:
\begin{enumerate}
\item Analyze the performance of our Artificial Intelligence approach for the {\bf synthesis} of programs up to: {\bf 10 program lines}, {\bf 10 observable variables} with binary domain and, an instruction set that comprises {\bf 10 instructions}. 
\item Analyze the performance of our Artificial Intelligence approach for the {\bf validation} of programs up to: {\bf 15 program lines}, {\bf 15 observable variables} with binary domain and, an instruction set that comprises {\bf 15 instructions}. 
\end{enumerate}

@@

We designed a 24-month workplan plan for the {\bf development of a user-interactive program synthesizer} that works in two different modes:
\begin{enumerate}
\item {\em Synthesis mode}. Takes as input a set of test cases that specify the TDD programming task to solve and outputs a program source code that passes these tests with a bug-free guarantee.
\item {\em Validation mode}. In this setting the {\em program synthesizer} receives an additional input, specifying the source code of the program to validate, and it outputs a {\em validation certificate} guaranteeing that the input program either succeeds (or fails) to solve each test case from the TDD programming task. 
\end{enumerate}

Figure~\ref{fig:gantt} details the proposed 24-month timeline for the development of the project. A deliverable is provided at the end of each task.

@@

  \item {\bf T2. Development of the system architecture (months 7-12})
    \begin{small}
      \begin{enumerate}
      \item The programming-into-planning compiler ({\em Compiler 1}). This system component parses the {\em TDD programming task} and produces an {\em AI Planning problem} encoded in the standard planning language PDDL.
      \item The plan-into-program compiler ({\em Compiler 2}). This system component extracts the program code and the corresponding validation certificate from the solution plan produced by an off-the-shelf AI planner.
      \end{enumerate}
\end{small}

@@

\item {\bf T3. Experiments and dissemination of results (months 13-24}).
   \begin{small}
      \begin{enumerate}
      \item Reporting the experimental performance of our AI approach for solving diverse TDD programming tasks. This task will follow an iterative workflow over the following subtasks:
      \begin{enumerate}
      \item Executing the system architecture in the {\em theoretical benchmarks} described in Section~\ref{sec:evaluation}.
      \item Analysis and validation of the obtained results.
      \item Tuning and repairing the system components, evaluation metrics and benchmarks according to the obtained results.                 
      \item Executing the system architecture in the {\em real-world benchmarks} introduced in Section~\ref{sec:evaluation}.
      \item Analysis and validation of the obtained results.         
      \item Tuning and repairing the system components, evaluation metrics and benchmarks according to the obtained results.                 
      \end{enumerate}
      \item Dissemination of the obtained theoretical and empirical results by submitting papers to top international conferences and journals in AI.        
      \end{enumerate}
\end{small}        
{\small{\bf\em  Deliverable T3:} Final report with obtained conclusions and produced publications.}
\end{enumerate}

\subsection{Presupuesto} 
Esta sección presenta la descripción detallada del presupuesto, de dos anualidades, para el desarrollo del proyecto:

\begin{table}[hbt!]
\begin{small}  
  \begin{tabular}{clc|r}
    {\large\bf Prioridad} & {\large\bf Descripcion} & {\large\bf Año} & {\large\bf Euros} \\\hline
1 & {\scriptsize Difusión de las actividades del grupo} & 1 & 4,000\\    
2 & {\scriptsize Viajes, manutención y alojamiento grupo de investigación} & 1 & 2,000\\
3 & {\scriptsize Viajes, manutención, alojamiento y ponencias investigadores invitados} & 1 & 2,000\\\hline
1 & {\scriptsize Difusión de las actividades del grupo} & 2 & 4,000\\    
2 & {\scriptsize Viajes, manutención y alojamiento grupo de investigación} & 2 & 2,000\\
3 & {\scriptsize Viajes, manutención, alojamiento y ponencias investigadores invitados} & 2 & 2,000\\\hline
\multicolumn{3}{l|}{} & 16,000 \\
  \end{tabular}
\end{small}
\caption{\small Presupuesto de dos anualidades para el desarrollo del proyecto.}
\end{table}

\begin{enumerate}
\item {\em Difusión de las actividades del grupo}. Esta partida contempla las inscripciones a congresos internacionales en Inteligencia Artificial, previsiblemente los tres siguientes: The International Conference on Automated Planning and Scheduling (h5-index 27). The International Joint Conference on Artificial Intelligence (h5-index 61). The AAAI Conference on Artificial Intelligence (h5-index 69). The International Conference on Machine Learning (h5-index 113). Todos los h5-index calculados de acuerdo a Google Scholar.
\item {\em Viajes, manutención y alojamiento grupo de investigación}. Esta partida contempla los gastos generados por la asistencia a los congresos internacionales mencionados anteriormente y la visita a grupos de investigacion relacionados con los objetivos del proyecto.
\item {\em Viajes, manutención, alojamiento y ponencias investigadores invitados}. Sergio Jimenez, PI del presente proyecto, mantiene un activa colaboración con el grupo {\em A.I. and Autonomy Lab} de la universidad de Melbourne {\tt\small\url{https://cis.unimelb.edu.au/agentlab}} asi como con el {\em Artificial Intelligence and Machine Learning Research group} ({\tt\small\url{https://www.upf.edu/web/ai-ml}}) de la Universitat Pompeu Fabra de Barcelona.  Esta partida contempla los gastos generados para el desarrollo de actividades conjuntas en el ámbito de la Comunitat Valenciana. 
\end{enumerate}

\section{Expected benefits of the research.}
\label{subsec:beneficios}
{\em AI Planning} has recently shown successful in {\em program testing} to generate {\em attack plans} that completed non-trivial software security tests~\cite{hoffmann2015simulated,steinmetz2016revisiting,shmaryahu2016constructing,steinmetz2016goal}. Promising research opportunities come from the application of {\em AI Planning} to {\em program synthesis} given that, {\em program synthesis} with test cases, can be seen as the {\em program testing} dual. In fact, our current work on {\em program synthesis} with AIP already produced several {\bf publications at the top international conferences and journals on Artificial Intelligence}~\cite{segovia2017generating,sergio:aprogramingb:ijcai16,sergio:aprograming:ijcai16,sergio:aprograming:icaps16,segovia:FSC:JAIR2018,segovia:programs:AIJ19} and is the recipient of the {\it 2016 distinguished paper award} at the International Joint Conference on Artificial Intelligence, the main international conference on {\em Artificial intelligence}.

The main benefit of this project is to {\bf provide new insights into the current understanding of how AI can assist programmers in the software development}.  Automated program synthesis promises to produce huge cost savings in comparison to conventional software development and to increase the quality, reliability and modifiability of programs.  In more detail, the expected benefits for this particular research project are four-fold:
\begin{enumerate}
\item A new {\bf evaluation methodology for the {\em Synthesis and Validation of Programs}}. The application of the exiting {\em AI Planning} technology to program synthesis can provide new evaluation metrics that assess how well a program covers a set of {\em input-output} test cases.  
\item An empirical {\bf study on the performance of the state-of-the-art AI planners for the {\em Synthesis and Validation of Programs}}. Research in AI algorithms is too often tested with laboratory problems and AIP is not an exception. Most of the new planning algorithms are only tested within the benchmarks of the International Planing Competition~\cite{vallati:IPC:AI15}. This project will help to meet the computational and expressiveness limits of AI planners when addressing real-world programming tasks. 
\item The {\bf development of open software and open benchmarks for the {\em Synthesis and Validation of Programs}}. We strongly belief that reproducibility and open knowledge are essential to the advance of the research on computer science. With this regard, we plan to develop a {\em github} repository where we make available the developed source code and benchmarks.
\item {\bf International dissemination of the obtained scientific results}. The scientific results obtained during the development of the project will be submitted to top Artificial Intelligence conferences (such as IJCAI, AAAI, ICML and ICAPS) and to the main journals in the AI field (such as AIJ, JMLR and JAIR).
\end{enumerate}
